---
title: "Additional Practice: Week 4 — Data Exploration"
output: pdf_document
---

```{r}
library(tidyverse)
```

This week, we'll explore stomach cancer mortality rates by county using a dataset from the Institute for Health Metrics and Evaluation [(Source)](https://ghdx.healthdata.org/record/ihme-data/united-states-stomach-cancer-mortality-by-county-race-ethnicity-2000-2019).

Load the `stomach_cancer_deaths_california.csv` dataset from the [course GitHub repo](https://github.com/r-pophealth-research/online-course/tree/main/data) in the cell below

```{r}
stomach_cancer_deaths = read_csv("https://raw.githubusercontent.com/r-pophealth-research/online-course/main/data/stomach_cancer_deaths_california.csv")
```

# Part 1: Data Summaries

Start by describing the structure of the data frame.

How many rows and columns are there?

```{r}

```

What variables are in the dataset? 

```{r}

```

How many missing values (`NA`s) are there in the `val` column? (Hint: use `is.na()` to check if a value is equal to `NA`)

```{r}

```

What are the unique values of sex, age, and race?

```{r}
...
```

# Part 2: Visualizing Data

Now that you're familiar with the data, make some plots to visualize the distribution of death rates among different subsets of the population. 

First, make a histogram that shows the distribution of all-race, age-standardized death rates. 

```{r}
total_rates = 
  stomach_cancer_deaths %>% 
    filter(race_name == ..., age_name == ...) %>% 
    select(location_name, val, upper, lower)

hist(total_rates %>% pull(val), xlab = "Death Rates (per 100,000)", main = "Distribution of Mortality Rates")
```

Now make a histogram that shows the distribution of API, age-standardized death rates. 

```{r}
api_deaths = 
  stomach_cancer_deaths %>% 
  ...

...

```


# Part 3: Summary Tables

Using `group_by` and `summarize`, create a table that contains the number of measurements, mean mortality rate, and median mortality rate for each race in the dataset. Restrict the original data set to non-missing, age-standardize values. Try different ways of sorting/arranging your final table.

```{r}
stomach_cancer_deaths %>% 
  filter(age_name == "Age-standardized", 
         !is.na(val)) %>% 
  group_by(...) %>% 
  summarize(n_values = ...,
            mean_mortality = ..., 
            median_mortality = ...) %>% 
  arrange(median_mortality)
```

# Challenge

Write a function that plots age-standardized deaths rates for a given race. Use this function to plot death rates for the "Total" and "API" race groups, and confirm that it yields the same plots as above.

```{r}
plot_mortality_by_race = function(race_str) {
  ...
}

plot_mortality_by_race("Total")
plot_mortality_by_race("API")
```


```{r}
# Get unique values of race using R commands (don't just type them out!)
unique_race_values = ...

for (race in unique_race_values){
  ...
}
```

