---
title: "Additional Practice: Week 3 — Functions & Packages"
output: pdf_document
---

# Review

**if-else statement**

```         
if(condition){
  action
} else {
  action
}
```

**for loop**

```         
for(item in items){
  action_with_item
}
```

**function**

```         
name <- function(input){
  action_with_input
}
```

Notice a pattern here? The `if`, `for`, and `function` statements all have a similar format! In general, **when we are writing R code that's run in a special way (run only sometimes or run repeatedly)**, we provide some type of instructions in parentheses `()`. Then we provide the code we want to run based on those instructions in curly brackets `{}`. We need those curly brackets `{}` to tell R what code we want run in the special way and what code is just run normally.

Note that `for` and `function` both allow you to create new variables inside the parentheses `()` that you can use later inside the curly brackets `{}`.

Want some more practice with these? Check out [while loops](https://www.w3schools.com/r/r_while_loop.asp) which use the same format!

# Functions, Ifs, and Loops

Functions are a way to do the same action to different inputs.

Functions have 2 important parts:

1.  The part that goes in parentheses is the input. Not all functions have inputs, and that's okay.
2.  The part that goes in the curly brackets is the the action that you want to run each time.

You can then call on a function using its name and putting the input inside parentheses.

## Practice 1 - Simple Function

Write a simple function that takes a character input and prints (use `print()`) that input.

```{r}
# Write your function here
printer

# Check your function
printer(input = "Hello world")
printer(input = "Go trees!")
```

## Practice 2 - Multiple Inputs

Write a function that takes two inputs, `base` and `exponent`, that returns the base raised to the exponent.

```{r}
# Fill in the function with the inputs and code needed to calculate
exponentiate <- function(...) {

}

# Check your function
exponentiate(base=2, exponent=5) # Should be 32
exponentiate(base=5, exponent=-1) # Should be 0.2
```

## Practice 3 - Function + If

When using the built-in `log` function, taking the log of 0 outputs `-Inf` (negative infinity), and taking the log of a negative number outputs `NaN` (not a number).

```{r}
log(0)
log(-2)
```

Write a function called `smart_log` that takes any number as the input. If that number is zero or negative, return `NA`. Otherwise, return the log of that number.

```{r}
smart_log <- function(...){
  # If statement to check if the number is zero or negative that returns NA
  
  # Otherwise, return the log of the number
  
}

# Check your function
smart_log(num=10)
smart_log(num=0)
smart_log(num=-2)
```

## Practice 4 - Function + Loop

Now that we have our function, we can use a for-loop to find the log of a bunch of numbers.

Write a for-loop that calculates and and saves the log of each number in a vector.

```{r}
# Calculate the log of these numbers using the function you made above
# Save the results to logs
numbers <- c(1, -10, 20, 50, 0)
logs <- c()

# Write your for-loop that calculates and saves the log of each number using 
# the function you made above
for(...) {
  # Use the smart_log function to get the log and save it to a variable
  
  # Add that variable to the logs list
  
}

# Check your results
logs
```

## Practice 5 - Challenge: Complex function

Remember this formula?

$$
x=\frac{-b\pm\sqrt{b^2-4ac}}{2a}
$$

It's the quadratic formula from algebra, which tells us where a parabola intercepts the $x$ axis. Write a function that takes three values, `a`, `b`, and `c`, and returns a vector with two values - the two places where the parabola crosses the $x$ axis. But, if `a` is zero, return a vector with two `NA` values because that is not a parabola.

Hint: There is no $\pm$ in R. You'll have to do those separately.

```{r}
# Write your function
get_parab_x_intercepts

# Testing
get_parab_x_intercepts(a=2, b=4, c=-3) # Should be 0.5811 and -2.5811
get_parab_x_intercepts(a=-5, b=1, c=6) # Should be -1 and 1.2
get_parab_x_intercepts(a=0, b=10, c=2) # Should be NA and NA
```

# Get Help with Functions

R has a bunch of functions that have already been written and documented for us. For example `min()` and `max()`:

```{r}
min(1,2,3,4,5)
max(1,2,3,4,5)
```

These functions have documentation to tell us how to use them. You access the documentation using the `?`.

```{r}
?min()
?max()
```

You can also use your Tab key inside the parentheses to get a suggestion of what should go next. Note the `na.rm` argument appears with a brief description.

```{r}
min()
```

## Practice 6 - Sequences

Use the `?` operator to figure out how to use the `seq()` function get every fifth number from 0 to 90 (i.e. 0, 5, 10, ..., 90). Use the same function to get every 0.2 from 2 to 5 (i.e. 2.0, 2.2, 2.4, ..., 5).

```{r}

```

# Working with Data

## Import Data

When you download data (or get it from collaborators), it's best practice to store it in a `data` folder when working locally. The main way you will load CSV files is with `read_csv()`. Note, you need to load the `tidyverse` package before you use `read_csv()`.

Load `employment_data.csv` from the [course GitHub repo](https://github.com/r-pophealth-research/online-course/tree/main/data) and save it to a variable.

```{r}
library(tidyverse)

# employment_data <- read_csv("https://raw.githubusercontent.com/r-pophealth-research/online-course/main/data/employment_data.csv")
```

If you ever need to import data from Excel, you can install and load the `readxl` package. This package contains the `read_excel()` function that you can use just like `read_csv()`.

## Export Data

Once you have your data cleaned, you'll want to save your data set. You can save your data using the function `write_csv()`. This function needs two things: 1) a data set and 2) a file location to put the data set. The example below works in RStudio on your own computer.

```{r}
# write_csv(employment_data, file = "data/employment_data_copy.csv")
```

# Fun with Dates

Dates are notoriously hard to deal with. There are leap years and [leap seconds](https://www.youtube.com/watch?v=lv3KAvP--ZM). Some months have 30 days, some months have 31, and other months can't decide how many days they want to have. Also, there are a bunch of different ways of writing the same date.

## Lubridate Intro

Lubridate to the rescue! `lubridate` is one of the packages included in the `tidyverse`. For additional help getting started, you can check out the [Lubridate Cheat Sheet](https://rstudio.github.io/cheatsheets/lubridate.pdf). To create a date in `lubridate`, you will use one of the parsing functions. These functions take in strings and create dates.

```{r}
library(lubridate)

# Year, month, day, hours, minutes, seconds
ymd_hms("2024-01-02 03:04:05")

# Month, day, year, hours, minutes, seconds
mdy_hms("10/31/22 23:59:59")

# Year, month, day
ymd("2020-03-15")

# Month, day, year
mdy("1/24/24")

# It knows what days exist and what don't
ymd("2024-02-29")
ymd("2023-02-29")

# It has special functions to get the current date/time
today()
now()
```

Now you use the right functions to turn these written out dates into Lubridate dates. Use the cheat sheet to help!

```{r}
ymd("2024-01-01")
mdy("01-31-24")
dmy("31/01/24")
ymd_hm("2022-31-08 11:00")

# You can tranform these dates all at once!
mdy(c("11/08/16", "11/02/20", "11/05/24"))
```

## Employment Exercise

Our goal is to use the `employment_data` dataset you imported above to calculate the employee's total compensation over the time they've worked at The Company(TM).

### Practice 7 - Get the start date and end dates as Lubridates

Use the dataset and what you just learned to pull out the start dates and end dates and transform them into Lubridates. Put the start dates into a variable called `start_dates` and the end dates into a variable called `end_dates`.

```{r}

```

### Practice 8 - Challenge: Fix the end dates

If people don't have an end date, they're still working at The Company(TM). So we need to replace all of the `NA`s in `end_date` with today's date.

Just like how we used numbers and square brackets to find specific places in our data, we can use logicals in square brackets to do the same thing

```{r}
example_data <- c("a","b","c","d","e")
# Both of these lines get only the second element
example_data[2]
example_data[c(F, T, F, F, F)]
# Both of these lines get the thrid through the fifth element
example_data[3:5]
example_data[c(F, F, T, T, T)]
```

Using this knowledge, set all of the `NA`'s in end_date to today's date.

```{r}
# Hint
is.na(end_date)
end_date


```

### Practice 9 - Challenge: Figure out how long each person has been working here

Lubridate allows us to subtract dates just like any other number. Use this to calculate how long each employee has been working at The Company(TM).

```{r}

```

Lubridate also has a concept called an interval which stores the time between two dates accounting for leap years and everything else. Using `interval()` or `%--%`, create a vector of intervals that describes when each person works here.

```{r}

```

### Practice 10 - Challenge: Multiply the number of months the person has been working by their monthly salary

Note: The Company(TM) pays its employees the same amount regardless of how long a month is (because it's cheaper).

Lubridate allows us to divide dates like they're math. These functions include dividing an entire time span by another time span. For example:

```{r}
all_of_jan <- interval(start=mdy("01-01-2025"), end=mdy("01-31-2025"))
all_of_jan / weeks(1)
```

Now take the intervals you made and get the number of whole months each person worked. Then multiply that by their monthly salary.

```{r}

```
