---
title: "Additional Practice: Weeks 8 & 9 — Data Visualizations"
output: html_notebook
---

```{r}
library(tidyverse)
library(lubridate)
```

# Part 1: Data Exploration

Run the cell below to load the data

```{r}
# washb_df = read_csv("https://raw.githubusercontent.com/r-pophealth-research/online-course/main/data/washb_subset.csv")
# head(washb_df)
```

Use the cells below to briefly explore the data.

What are the treatment (`tr`) groups? How many measurements are in each treatment group?

```{r}

```

What date range does the study cover?

```{r}

```

What are the unique values of `season`?

```{r}

```

# Part 2: Data Processing

## Recoding Variables

For this analysis, let's consider all interventions as a single pooled group. Use the cell below to create a `pooled_tr` variable in `washb_df` that is equal to "Any WASH Intervention" if `tr != "Control"` and "Control" if `tr == "Control"`

```{r}
washb_df = washb_df %>% 
  mutate(pooled_tr = ...)

# Check that the only values in pooled_tr are "Control" and "Any WASH Intervention"
washb_df %>% pull(pooled_tr) %>% unique
```

We can also give the `season` variable more informative values. In this dataset, if season is "yes" that means the measurement was taken in the rainy season. If season is "no", the measurement was taken in the dry season. **Replace the values in `season`** with "Rainy Season" or "Dry Season" based on the current values of `season`

```{r}
washb_df = washb_df %>% 
  mutate(season = ...)

# Check that the values in season are "Rainy Season" and "Dry Season"
washb_df %>% pull(season) %>% unique
```

## Estimating Prevalence

Run the cell below to estimate the prevalence of diarrhea (`diar7d`) in the pooled vs intervention group.

```{r}
washb_df %>% 
  group_by(pooled_tr) %>% 
  summarize(prev = mean(diar7d, na.rm = T) * 100) # multiply by 100 to %
```

We're interested in whether there are seasonal differences in prevalence between groups. Try calculating the prevalence by both treatment group and season. What do you notice?

```{r}
washb_df %>% 
  group_by(...) %>% 
  summarize(...)
```

It looks like there might be some differences in intervention effectiveness by season. In other words, the difference in diarrhea prevalence between treatment arm is higher in the rainy season vs the dry season.

We can investigate this further by plotting diarrhea prevalence in each group **over time**. However, there are only a few measurements that are taken on a given day, so it might be more informative to look at **monthly prevalence.**

In the cell below, create a data frame called `prev_monthly` that calculates the prevalence of diarrhea in each group for every month.

```{r}
# Creates a copy of washb_df called washb_df_monthly, with a variable "month"
# The function floor_date can be used to "round" dates to the same month, as if 
# measurements were taken on the first of the month
washb_df_monthly = washb_df %>% 
  mutate(month = floor_date(date, "month"))

# Create a data frame with monthly prevalence by arm (pooled vs control)
prev_monthly = washb_df_monthly %>% 
  group_by(...) %>% 
  summarize(...)

prev_monthly
```

## Handling Missing Values

As you scroll through the monthly prevalence, you might notice that there are some months where there were no measurements (and therefore no prevalence estimates).

Run the cell below to create a bar plot that shows the number of measurements in each month. Ideally, we'll see two estimates (one for each treatment group) in each month.

```{r}
ggplot(prev_monthly, aes(x = month)) + 
  geom_bar()
```

There seems to be a gap in measurements in the middle of the date range! When we go to plot trends over time, a line will always connect neighboring points even if there's a gap in between. For a visual of this, **fill in the cell below to** **make a line plot of monthly diarrhea prevalence *in the control group* over time.**

```{r}
control_monthly_prev = prev_monthly %>% filter(pooled_tr == ...)

ggplot(control_monthly_prev, aes(x = ..., y = ...)) + 
  geom_line()
```

In this plot, prevalence estimates from Sept 2014 will connect to those from Dec 2014 **even though we don't have data for October or November**. This can be misleading since the line suggests that a particular trend, even though we don't have the data to support it.

To prevent this, we can add rows to the data frame for the "missing" months where the prevalence estimates are "NA". Later, this will tell ggplot not to plot anything for those months. Run the code below, and read the comments to make sure you understand what each line is doing.

```{r}
# Creates a data frame with all possible combinations of months and tr groups
#   seq.Date(...) finds all months in the study period 
#   unique(...) finds all unique treatment groups in the study 
all_dates = expand.grid(month = seq.Date(from = min(prev_monthly$month),
                                         to = max(prev_monthly$month),
                                         by = "month"),
                        pooled_tr = unique(prev_monthly$pooled_tr))

# Left join to keep all months, match prevalence estimates if they exist or fill NA if they don't
all_prev_monthly = all_dates %>% left_join(prev_monthly, by = c("month", "pooled_tr"))

# Check that "missing" months are present in the new dataset, with prev values of NA
all_prev_monthly %>% filter(is.na(prev))
```

# Part 3: Plotting!

We're ready to make a figure! First, plot the monthly prevalence of diarrhea in the **control** group over time using values from `all_prev_monthly`.

```{r}
control_all_prev_monthly = all_prev_monthly %>% filter(pooled_tr == ...)

ggplot(...) + 
  ...
```

Now, plot the monthly prevalence of diarrhea in the **pooled intervention** group over time using values from `all_prev_monthly`.

```{r}
pooled_tr_all_prev_monthly = all_prev_monthly %>% 
  filter(pooled_tr == "Any WASH Intervention")

ggplot(...) + 
  ...
```

We now have a visual of how diarrhea prevalence changes over time in each, but it's hard to compare how different these trends are when the lines are on separate plots...

Let's try to use the **group** aesthetic to show both lines on the same plot.

```{r}
ggplot(..., aes(..., group = ...)) + 
  ...
```

Now, we can see that there's a clear difference between the lines! But which one is which?? Let's tell ggplot to **vary the color of the lines by treatment group** too.

```{r}
ggplot(..., aes(..., group = ..., color = ...)) + 
  ...
```

Now that we can distinguish between the lines, this figure suggest that there are larger seasonal spikes in diarrhea in the control group vs the intervention groups (monsoon season in Bangladesh is around June-October)

**Challenge:** Try customizing this plot! You're free to explore anything you want, but consider formatting the axis labels, using a custom color palette, playing around with the `theme_XX()` options, or moving the position of the legend.

```{r}

```

# Challenge: Plotting prevalence ratio estimates

We've made a nice descriptive plot above, but we can make a stronger argument about the impact of weather by conducting a formal statistical analysis and plotting our estimates of intervention effects.

Since we're looking at diarrhea prevalence, we can compare prevalence ratios associated with the intervention during the rainy season vs during the dry season.

Prevalence ratios can be measured using log-binomial regression [[Reference](https://www.bookdown.org/rwnahhas/RMPH/blr-log-binomial.html)]. The prevalence ratio + 95% CIs can be calculated by taking the exponent of the coefficients associated with treatment in your model.

Run the code below to generate the prevalence ratio estimates and CIs for measurements taken the rainy season. Read the comments to understand what each chunk of code does.

```{r}
# Resets the levels of tr so "Control" is the reference group
washb_df = washb_df %>% 
  mutate(tr = factor(tr, levels = c("Control", "Combined WSH", 
                                    "Water", "Sanitation", "Handwashing")))

# Get a log-binomial model to estimate prevalence ratios
rainy_season_model = glm(diar7d ~ tr, 
                         family = binomial(link = "log"), 
                         data = washb_df %>% filter(season == "Rainy Season"))

# Extract coefficient, exponentiate to find prevalence ratios 
# ([-1] excludes the intercept)
rainy_season_prs = exp(coef(rainy_season_model)[-1])
rainy_season_prs

# Extract confidence intervals, exponentiate to find prevalence ratios 
rainy_season_pr_cis = exp(confint(rainy_season_model)[-1, ])
rainy_season_pr_cis

# Combine PR estimates and CIs into a single table using `bind_cols`
combined_rainy_season_estimates = 
  data.frame(pr_est = rainy_season_prs) %>% 
  bind_cols(rainy_season_pr_cis)

# Formatting and labeling for estimates table. 
combined_rainy_season_estimates = combined_rainy_season_estimates %>% 
  rownames_to_column("tr") %>% # Create treatment column from row names
  mutate(tr = str_remove(tr, "tr"), # Remove "tr" default model output
         season = "Rainy season") %>%  # Indicate that estimates are from the rainy season
  rename("ci_lb" = "2.5 %", "ci_ub" = "97.5 %") #Rename CI columns

combined_rainy_season_estimates
```

Adapt the code above to generate a table called `combined_dry_season_estimates` that contains the estimates and 95% CI for PRs in each treatment group during the dry period (season == "Dry Season"). Make sure it has the *exact* same column names as `combined_rainy_season_estimates` (tr, pr_est, ci_lb, ci_ub, season).

```{r}
# Get a log-binomial model to estimate prevalence ratios
...

# Extract coefficient, exponentiate to find prevalence ratios 
# ([-1] excludes the intercept)
...

# Extract confidence intervals, exponentiate to find prevalence ratios 
...

# Combine PR estimates and CIs into a single table using `bind_cols`
...

# Formatting and labeling for estimates table. 
...
```

Run the cell below to merge the estimates + CIs from the rainy and dry season into a data frame called `combined_pr_estimates`.

```{r}
combined_pr_estimates = bind_rows(combined_rainy_season_estimates,
                                  combined_dry_season_estimates)
```

Use `ggplot` to plot the estimates and error bars contained in `combined_pr_estimates`. You may need to use `position_dodge()` in `geom_point()` and `geom_line()` to avoid overlap in estimates by season.

Hint: use `group = season` and `color = season`.

```{r}
ggplot(...) + 
  ...
```
