---
title: "Additional Practice: Week 5 — Data Manipulation"
output: html_document
---

```{r}
library(tidyverse)
```

This week, we'll explore dengue infections using a dataset from [this study published in *Nature Microbiology*](https://www.nature.com/articles/s41564-023-01543-3 "Household immunity and individual risk of infection with dengue virus in a prospective, longitudinal cohort study").

Load the `dengue_infection_data.csv` dataset from the [course GitHub repo](https://github.com/r-pophealth-research/online-course/tree/main/data) in the cell below.

In this dataset, each row captures the number of infections reported for each individual during the study. Each individual is assigned an `ID`, and individuals who share a household share a `hh_ID`.

```{r}
# dengue_infections = read_csv("https://raw.githubusercontent.com/r-pophealth-research/online-course/main/data/dengue_infection_data.csv")
```

# Part 1: Data Exploration

Start by exploring the data. We've provided some suggested steps, but feel free to add more cells if there's anything you want to take a closer look at.

What the dimensions (number of rows, number of columns) of the dataset?

```{r}

```

Summarize the dataset using `summary()`. What are the columns in the data? What do they represent?

```{r}

```

How many individuals are captured in the dataset? How many households? In other words, how many unique values of the `ID` and `hh_ID` columns are there?

```{r}

```

Make a histogram of the number of asymptomatic cases in the dataset (`n_asymp_infections`)

```{r}

```

**Challenge:** Estimate the average size of households in the dataset.

```{r}

```

# Part 2: Creating and Modifying Variables

Run the cell below to create a copy of `dengue_infections` called `dengue_infections_clean`. We will be using `dengue_infections_clean` for the rest of the notebook, where we'll be adding new variables + modifying existing variables. Generally, it's good practice to keep your original "untouched", and make modifications on a copy.

```{r}
dengue_infections_clean = dengue_infections
```

Suppose we find out that ages were not properly recorded for individuals with IDs 4, 87, and 134. We want to retain their infection information for future analyses, but don't want the incorrect ages to bias any results. Modify the `age` column so the ages for IDs 4, 87, and 134 are equal to `NA`, but that `age` is unchanged for all other IDs.

```{r}
dengue_infections_clean = dengue_infections_clean %>% 
  mutate(... = if_else(ID %in% c(4, 87, 134), ..., ...))
```

Create a column in `dengue_infections_clean` called `any_infection`, which is equal to `1` if the individual had experienced either a symptomatic or asymptomatic infection during the trial and `0` otherwise.

```{r}
dengue_infections_clean = dengue_infections_clean %>% 
  ...
```

Create a column in `dengue_infections_clean` called `age_group`, which is equal to "\<18" if the individual is between the ages of 0 and 17 years old, "18-64" if the individual is between the ages of 18 and 64 years old, and 65+ if the individual is 65 or more years old.

```{r}
dengue_infections_clean = dengue_infections_clean %>% 
  ...
```

# Part 3: Joining Data Sets

We also have access to household-level characteristics in `dengue_hh_data.csv`. Run the cell below to load that data.

```{r}
# dengue_hh_data = read_csv("https://raw.githubusercontent.com/r-pophealth-research/online-course/main/data/dengue_hh_data.csv")
```

Join the *cleaned* individual and household level data to create a dataset called `merged_dengue_data`, where there is still only one row per individual

```{r}
merged_dengue_data = ...
```

To confirm that you've completed the join correctly, run the cell below and make sure it returns `TRUE`.

```{r}
nrow(merged_dengue_data) == nrow(dengue_infections)
```

# Bonus: Grouping and Reshaping Data

Suppose we were interested in differences in dengue prevalence by age. Use `group_by` and `summarize` to estimate the count and prevalence of any dengue (both symptomatic and asymptomatic) in each age group. Use the grouped data to get the percent of \<18 with dengue divided by the percent of 18-65 with dengue.

```{r}
prev_by_age = ...

prev_under18 = ...
prev_18to65 = ...

# Prevalence ratio of <18 vs 18-65
prev_under18 / prev_18to65
```

To more easily compare symptomatic and asymptomatic infections, we might want to restructure our data from having seperate `n_symp_infections` + `n_asymp_infections` columns to a `infection_type` (with the values "symp" or "asymp") columns + a `n_infections` (with the number of infections) column. Use `pivot_longer` to transform your data in this way.

```{r}
long_dengue_df = ...

long_dengue_df
```
