---
title: "HW1_markdown_and_data_types"
output: pdf_document
---

# Homework 1

Welcome to your first homework! Edit the code cells below and run them to check your answers.

## Part 1: Markdown

### Question 1 (2 points)

Create a code block that calculates 1,248 squared.

## Part 2: Data Types

### Question 2 (3 points)

Replace the `#Your_code_here`'s with *appropriate data* based on the variable names. Hint: You will use 1 of each data type.

```{r}
average_height_of_a_girrafe_in_meters = #Your_code_here
the_best_city_in_the_world = #Your_code_here
is_strawberry_the_best_ice_cream_flavor = #Your_code_here
```

## Part 3: Comparisons

### Question 3 (2 points)

Write a comparison using numerics that evaluates to `TRUE`.

```{r}

```

### Question 4 (2 points)

Write a comparison using characters that evaluates to `FALSE`.

```{r}

```

### Question 5 (2 points)

Write a comparison that uses NA and, based on your experience, explain what happens anytime you use NA in a comparison. Hint: Try a bunch of comparisons using an NA and see what you get.

```{r}

```

Briefly explain what is happening:

## Challenge: Think About Data in Research

Search [PubMed](https://pubmed.ncbi.nlm.nih.gov/) for a public health topic you are interested in (ex. "opioid addiction"). In the left sidebar under "Article Type", click "Clinical Trial" to only view trials. Pick an article you find interesting, and briefly answer the questions below. If you're having trouble answering the questions, you can pick a different article.

*NOTE:* Some articles may require access through your institution's library to view the full text.

### Challenge 1 (1 point)

Please paste a link to the article here:

### Challenge 2 (1 points)

What was the study objective?

### Challenge 3 (4 points)

Describe in words and using R code at least two inclusion criteria for the study. Inclusion criteria are requirements described in the paper that participants had to meet before joining the study. Replace the examples with at least two of your own inclusion criteria from your research article.

-   Example: Patients 40 years old or greater - age \>= 40
-   Example: Patients with greater than 140 mm Hg blood pressure - blood_pressure \> 140
