---
title: "Homework 2"
output: pdf_document
---

This assignment covers materials from Weeks 2 and 3, including data structures (vectors and data frames) and functions. You should be able to complete Parts 1-2 after the Week 2 lecture and Part 3 after the Week 3 lecture. **You only need to complete the "Challenge Questions" if you have enrolled in the class for 2 units.**

If you have questions about this assignment, please attend office hours or create a discussion post on the course Ed.

### Quick note about style points

There are two points each homework given to "style" to give you credit for writing readable code. Readable code always has informative variable names like `blood_pressure`, `format_sentence`, or `height_boxplot`. Variable names like `data`, `x`, or `num` are almost never appropriate and are confusing in the long run. Even adding a prefix like `imported_data` allows for future variables to be named `cleaned_data` and `table_data` improving clarity. If you are writing longer, more complex code, use `# Comments` to describe what each part does. Comments are especially useful for before functions where you might describe what the function does. For example:

``` r
# Function to add two numbers together
# Accepts two numerics and returns the sum numeric
add_nums = function(x, y) {
  return(x+y)
}
```

# Part 1: Vectors

Suppose that we want to study the impacts of household income on health outcomes. We survey households in the community and collect the following data:

-   Household 1: \$35,294
-   Household 2: \$74,243
-   Household 3: \$93,866
-   Household 4: \$12,325
-   Household 5: \$53,243

## Question 1 (2 points)

Create a vector called `household_incomes` that contains the survey data, as numerics. Make sure that the first item in the vector corresponds to household 1, the second item corresponds to household 2, etc.

```{r}
household_incomes = ...
household_incomes
```

## Question 2 (2 points)

We continue to survey the community and collect incomes for 2 additional households. Household 6 reported an income of \$124,950 and Household 7 reported an income of \$42,738. Create a vector `updated_household_incomes` that extends the `household_incomes` vector with the two new income measurements

```{r}
updated_household_incomes = ...
updated_household_incomes
```

## Question 3 (2 point)

Calculate summary statistics for `updated_household_incomes`. Save `mean_income`, `median_income`, `min_income`, and `max_income` to their respective values.

```{r}
mean_income = ...
print(paste("Mean income =", mean_income))

median_income = ...
print(paste("Median income =", median_income))

min_income = ...
print(paste("Minimum income =", min_income))

max_income = ...
print(paste("Maximum income =", max_income))
```

## Question 4 (3 points)

We decide that we don't need to measure the exact dollar amount of these incomes. Instead, we're interested in rounding incomes to the nearest *thousands*. For example, \$35,294 should be 35 and \$93,866 should be 94 - note how the rounded values are whole numbers. Create a vector `rounded_incomes` that contain household incomes in thousands.

Hint 1: You will have to complete a vector operation on `updated_household_incomes`. Hint 2: The result of your vector operation may include many decimal values. You may need to use the `round()` function to gain whole numbers. You can run `?round` to learn how to use the function.

```{r}
rounded_incomes = ...
rounded_incomes
```

# Part 2: Data Frames

We're interested in studying hypertension (high blood pressure) in a patient population, and extract blood pressure values from their electronic health records. Blood pressure values consist of two values (ex 124/75), where the top value is the systolic number and the bottom value is the diastolic number.

## Question 5 (2 points)

We collect the following data:

-   Patient 1: 120/80 mmHg

-   Patient 2: 130/85 mmHg

-   Patient 3: 118/76 mmHg

-   Patient 4: 140/90 mmHg

-   Patient 5: 122/78 mmHg

-   Patient 6: 150/100 mmHg

-   Patient 7: 126/82 mmHg

Create a data frame `blood_pressure_df` that contains these data points. It should have three columns: `patient_id`, `systolic_value`, and `diastolic_value`.

```{r}
blood_pressure_df = data.frame(... = ...,
                               ... = ...,
                               ... = ...)
blood_pressure_df
```

## Question 6 (1 point)

Pull the row in `blood_pressure_df` that corresponds to the measurement for Patient 5 and save it to the variable `patient_5_df`

```{r}
patient_5_df = ...
patient_5_df
```

## Question 7 (1 point)

Pull systolic blood pressure values as a one-column data frame called `systolic_df`.

Hint: use `class()` to check that `systolic_df` is a `data.frame` type.

```{r}
systolic_df = ...
systolic_df
```

## Question 8 (1 point)

Pull systolic blood pressure values as a vector of numerics called `systolic_vector`.

Hint: use `class()` to check that `systolic_vector` is a `numeric` type.

```{r}
systolic_vector = ...
systolic_vector
```

# Part 3: Functions

## Question 9 (2 points)

Write a function that takes someone's systolic and diastolic blood pressure and returns "Normal" or "Hypertension", or "Severe Hypertension" based on the person's blood pressure. The American Heart Association defines **hypertension** as having a systolic blood pressure greater than or equal to 130 mmHg **OR** a diastolic blood pressure greater than or equal to 80 mmHg. **Severe hypertension** corresponds to systolic blood pressure of greater than or equal to 140 mmHg **OR** diastolic blood pressure greater than or equal 90 mmHg.

Apply your function to a patient with a systolic BP of 120 and diastolic BP of 75, and a patient with a systolic BP of 150 and diastolic BP of 85.

```{r}
classify_hypertension = ...

# Apply your function to a patient with systolic = 120, diastolic = 75
classify_hypertension(..., ...)

# Apply your function to a patient with systolic = 150, diastolic = 85
classify_hypertension(..., ...)
```

## Challenge Question 1 (3 points)

Write a for loop using the `classify_hypertension` function to create a vector called `hypertension_statuses` that stores the hypertension status of each person in the `blood_pressure_df` data frame.

Hint: You can use the for loop to look at each row of the data frame using the row index.

```{r}
# Data frame to use
blood_pressure_df

# Initiate the hypertension_statuses vector
hypertension_statuses = c()

# Write for loop that will look at each row using the row index i
for(i in ...){
  # For each row, extract this patient's systolic and diastolic blood pressure
  current_sys_bp = blood_pressure_df[i,...]
  current_dia_bp = blood_pressure_df[i,...]
  
  # Implement your function to save the correct hypertension status for each patient
  # to the hypertension_statuses vector
  current_hypertension_status = ...
  hypertension_statuses = ...
}

# Show the contents of the hypertension_statuses vector
hypertension_statuses
```

## Challenge Question 2 (2 points)

Don't like writing a for loop each time you want to do work on a data frame? Neither do we. R gives us the function `mapply()` (multiple apply) which passes values in order to a function. Use `?mapply` and Google to figure out how to create a vector with the hypertension status of each patient using your function, the patient data frame `blood_pressure_df`, and `mapply`.

```{r}
...
```

# Submission Instructions

Knit this R notebook as a PDF and submit it to the corresponding Gradescope assignment. Check your output to make sure that all your code and its output is displayed.

Once you successfully submit, you will be asked to match each question to its respective page on your submission. **Do not skip this step.**
