---
title: "Homework 3"
output: pdf_document
---

# Homework 3

This assignment covers materials from Weeks 4 and 5, including data exploration and data manipulation. You should be able to complete Part 1 after the Week 4 lecture and Parts 2-3 after the Week 5 lecture. **You only need to complete the "Challenge Questions" if you have enrolled in the class for 2 units.**

In each question you are welcome to use additional lines of code or create intermediate variables, but be sure to save your final answer to the specified variable.

If you have questions about this assignment, please attend office hours or create a discussion post on the course Ed.

# Loading packages and data

Run the cell below to load the required packages

```{r}
library(tidyverse)
library(here)
```

## Data Overview 

In this homework, we'll be working with hospitalization data collected by the California Department of Public Health. Specifically, we'll be looking at the number of COVID-19 and influenza hospitalizations in California during 2021. **Download the `covid19_hospitalizations.csv` and `influenza_hospitalizations.csv` files from Canvas and save them in the data folder of your class directory**

The first dataset, loaded as `covid19_hospitalizations` contains daily data on the number of hospital patients with COVID-19, by county. It contains the following columns:
* `county`: a string, the name of a county
* `todays_date`: a date, the day that counts were recorded for
* `hospitalized_covid_confirmed_patients`/`hospitalized_covid_suspected_patients`: a numeric, the number of patients in the hospital (including the ICU) that have a lab-confirmed/suspected case of COVID-19 
* `icu_covid_confirmed_patients`/`icu_covid_suspected_patients`: a numeric, the number of patients in the ICU that have a lab-confirmed/suspected case of COVID-19
* `all_hospital_beds` / `icu_available_beds`: a numeric, the total number of available hospital/ICU beds.

Run the cell below to load the COVID-19 hospitalization data. 

```{r}
#Source: https://data.chhs.ca.gov/dataset/covid-19-hospital-data
covid19_hospitalizations = read_csv(here("data", "covid19_hospitalizations.csv"))
head(covid19_hospitalizations)
```

The second dataset, loaded as `flu_hospitalizations`, contains daily data on the number of influenza hospitalizations across the whole state. It contains the following columns:
* `date`: a date, the day that counts were recorded for
* `patient_vol_influenza`: a numeric, the total number of inpatients with a influenza diagnosis
* `patient_vol_influenza_icu`: a numeric, the total number of ICU patients with a influenza diagnosis
* `admits_influenza`: a numeric, the total number of patients who were newly admitted and had a influenza diagnosis on the same day

Run the cell below to load the influenza hospitalization data.

```{r}
#Source: https://data.chhs.ca.gov/dataset/respiratory-virus-dashboard-metrics 
flu_hospitalizations = read_csv(here("data", "influenza_hospitalizations.csv")) 
head(flu_hospitalizations)
```

# Part 1: Exploring Data

## Question 1 (3 points)

On which day was there the most confirmed COVID-19 patients in the ICU in Alameda County? Use tidyverse commands to find the answer from the `covid19_hospitalizations` dataset, and save the date to the variable `highest_icu_date_alameda`. 

Hint: There are many ways you can approach this question. In one possible method, you can use some combination of `filter()`, `pull()`, `arrange()`

```{r}
highest_icu_date_alameda = ...
highest_icu_date_alameda
```

## Question 2 (2 points)

How many days had over 50 influenza hospitalizations? Use tidyverse commands to find the answer from the `flu_hospitalizations` dataset, and save the value to the variable `n_days_over50_flu_hosp`.

Hint: There are many ways you can approach this question. In one possible method, you can use some combination of `filter()` and `nrow()`. 

```{r}
n_days_over50_flu_hosp = ...
n_days_over50_flu_hosp
```

## Question 3 (3 points)

What was the maximum number of ICU patients with influenza between September 1st and December 31st? Use tidyverse commands to find the answer from the `flu_hospitalizations` dataset, and save the value to the variable `max_flu_icu_patients`.

Hint: you can use `filter` on dates but need to compare the date column to other `date` value, similar to how you have to compare numbers to numbers and not numbers to characters. You can convert a string of format YYYY-MM-DD to a date using `ymd`. For example, this code checks if a date is between March 14, 2018 and November 8, 2019: `filter(date >= ymd("2018-03-14"), date <= ymd("2019-11-08"))`.

```{r}
max_flu_icu_patients = ...
max_flu_icu_patients
```

## Question 4 (3 points)

Create a scatter plot comparing the number of inpatient hospitalizations and the number of new influenza admits. Place current hospitalizations on the x-axis and new admissions on the y-axis. Make sure to include descriptive axis labels and a plot title. 

```{r}
...
```

## Challenge Question 1 (2 points)

Create a vector called `county_covid_hosp` that contains the total number of confirmed COVID-19 hospitalizations for each county. Plot the distribution of the new vector as a histogram. Make sure to include descriptive axis labels and a plot title. 

```{r}
county_covid_hosp = ...
...
```

# Part 2: Creating Variables

## Question 5 (3 points)

Create a dataframe called `all_covid19_hospitalizations` that is a copy of `covid19_hospitalizations` with two new columns (1) `all_covid_hospitalizations`, which is a sum of the confirmed and suspected COVID-19 hospitalizations and (2) `all_covid_icu`, which is a sum of the confirmed and suspected COVID-19 ICU patients.

```{r}
all_covid19_hospitalizations = ... 
all_covid19_hospitalizations
```

## Question 6 (3 points)

Create a dataframe called `flu_admits_levels` that is a copy of `flu_hospitalizations` with a new column `admit_level` that has the value `low` when there are 0-5 new admits with influenza, `medium` where there are 6-10 new admits with influenza, and `high` if there are over 10 new admits with influenza.

```{r}
flu_admits_levels = ... 
flu_admits_levels
```

## Challenge Question 2 (4 points)

Create a dataframe called `covid_icu_capacity` that is a copy of `covid19_hospitalizations` with a new column called `prop_icu_beds_occupied`, which is the proportion of ICU beds that are currently occupied by patients with either confirmed or suspected COVID-19. 

Assume that the total number of ICU beds is the sum of available beds, beds taken by confirmed COVID-19 patients, and beds taken by suspected COVID-19 patients. Of course, this isn't realistic since it ignores any ICU beds taken by non-COVID patients, but we'll make the simplification for the sake of this question. 

Hint: you may need to make two intermediate calculations for the number of ICU beds occupied by COVID-19 patients and the number of total ICU beds before calculating `prop_icu_beds_occupied`.

```{r}
covid_icu_capacity = ...
covid_icu_capacity
```

# Part 3: Joining Data

## Question 7 (2 points)

The code below creates a dataframe `daily_covid19_hospitalizations` that adds up COVID-19 hospitalizations across all counties for each day of 2021.

```{r}
daily_covid19_hospitalizations = covid19_hospitalizations %>% 
  group_by(todays_date) %>% 
  summarize(hospitalized_covid_confirmed_patients = sum(hospitalized_covid_confirmed_patients), 
            hospitalized_covid_suspected_patients = sum(hospitalized_covid_suspected_patients), 
            icu_covid_confirmed_patients = sum(icu_covid_confirmed_patients), 
            icu_covid_suspected_patients = sum(icu_covid_suspected_patients))

head(daily_covid19_hospitalizations)
```

The code below shows the first few rows of `flu_hospitalizations` for your convenience. 

```{r}
head(flu_hospitalizations)
```

Merge `daily_covid19_hospitalizations` and `flu_hospitalizations` by date to create a new data frame called `respiratory_hospitalizations`. 

```{r}
respiratory_hospitalizations = ...
respiratory_hospitalizations
```

## Challenge Question 3 (3 points)

Using your newly created `respiratory_hospitalizations` dataframe, create a scatterplot comparing the total influenza non-ICU hospitalizations vs. COVID-19 non-ICU hospitalizations patients. Include both confirmed and suspected COVID-19 cases in your calculations. Your plot should have an informative title and axis labels. 

Hint: start by creating columns in your dataset called `total_covid_nonicu_hospitalizations` and `total_flu_nonicu_hospitalizations`.

```{r}
...
```

# Submission Instructions

Knit this R notebook as a PDF and submit it to the corresponding Gradescope assignment. Check the file to make sure that all your code and its output is displayed.

Once you successfully submit, you will be asked to match each question to its respective page on your submission. **Do not skip this step.**
