---
title: "Homework 5"
output: pdf_document
---

# Homework 5

This assignment covers materials on data visualizations from Weeks 8+9. **You only need to complete the "Challenge Questions" if you have enrolled in the class for 2 units.** If you have questions about this assignment, please attend office hours or create a discussion post on the course Ed.

For this homework, **we require you to generate all plots using `ggplot2`.** Figures generated using base R will only receive up to 50% of the points for each question.

In each question you are welcome to use additional lines of code or create intermediate variables, but be sure to save your final answer to the specified variable. **You will receive a style point penalty if your code is not indented correctly.** If you are not sure if your code is indented correctly, you can bring it back into one line and press "Enter". RStudio will add in the correct spacing for you.

# Loading packages and data

Run the cell below to load the required packages

```{r}
library(tidyverse)
library(here)
```

## Data Overview

In this homework, we'll be working indicators of noncommunicable diseases (NCD) like heart disease, cancer, and diabetes and associated risk factors collected by the WHO Global Health Observatory. **Download the `ncd_risk_factors.csv` file from Canvas and save them in the data folder of your class directory**

The dataset contains country-level annual estimates on the prevalence of different NCDs and risk factors that may contribute to these conditions. It contains the following columns:

-   `country`: a string, the name of a country

-   `year`: a numeric, the year that estimates were recorded

-   `indicator_abbr`: a character, the abbreviated code for the indicator

-   `indicator_name`: a character, the full name of the indicator

-   `sex`: a character, the total number of available hospital/ICU beds

-   `estimate`: a numeric, the point estimate of the corresponding indicator

-   `ci_lb`: a numeric, the lower bound of the confidence interval of the indicator esitmate

-   `ci_ub`: a numeric, the upper bound of the confidence interval of the indicator esitmate

Run the cell below to load the NCD data as `ncd_data`

```{r}
#Source: https://www.who.int/data/sets/health-inequality-monitor-dataset#gho
ncd_data = read_csv(here("data", "ncd_risk_factors.csv"))
head(ncd_data)
```

Run the cell below to see the mappings between each `indicator_abbr` and `indicator_name`. We can see that the data contains a variable `NCDMORT3070` that describes the probability of death from a chronic conditions between the ages of 30-70, as well as factors related to blood pressure, hypertension, BMI, and substance use.

```{r}
ncd_data %>% select(indicator_abbr, indicator_name) %>% distinct() 
```

# Part 1: Plotting Univariate and Bivariate Relationship

## Question 1 (3 points)

Create a histogram to plot the distribution of estimated prevalence of heavy episodic drinking for females in 2016. Make sure to add a title and informative axis labels to your plot.

```{r}
heavy_drinking_prev_df = ncd_data %>% filter(indicator_abbr == "SA_0000001739", sex == "Female")

ggplot(..., aes(...)) + 
  ...
```

## Question 2 (3 points)

Create a boxplot to compare the distributions of estimated prevalence of tobacco use in 2015 for males vs females. Make sure to add a title and informative axis labels to your plot.

```{r}
tobacco_use_df = ncd_data %>% 
  filter(indicator_abbr == "M_Est_tob_curr", year == 2015)

ggplot(..., aes(...)) + 
  ...
```

## Question 3 (3 points)

Create a scatterplot to visualize the relationship between insufficient physical activity (`NCD_PAA`) and probability of death from a NCD between the age of 30 and 70 (`NCDMORT3070`) for males in 2016. Place `NCD_PAA` on the x-axis and `NCDMORT3070` on the y-axis. Make sure to add a title and informative axis labels to your plot.

```{r}
bp_mortality_df = ncd_data %>% 
  filter(year == 2016, sex == "Male", 
         indicator_abbr %in% c("NCD_PAA", "NCDMORT3070")) %>% 
  pivot_wider(id_cols = country, 
              names_from = indicator_abbr,
              values_from = estimate)

ggplot(..., aes(...)) + 
  ...
```

# Part 2: Plotting Estimates and CIs

## Question 4 (4 points)

Plot the estimates and confidence intervals (using `geom_linerange()`) of probability of death from a NCD between the age of 30 and 70 (`NCDMORT3070`) for females in Canada, Mexico, and the United States in 2018. Make sure to add a title and informative axis labels to your plot.

```{r}
mortality_usa_can_mex_df = ncd_data %>% 
  filter(year == 2016, sex == "Female",
         indicator_abbr == "NCDMORT3070",
         country %in% c("United States of America", "Canada", "Mexico"))

ggplot(..., aes(...)) + 
  ... 
```

## Question 5 (4 points)

Using a line plot, visualize the estimates and confidence intervals (using `geom_ribbon()`) of probability of death from a NCD between the age of 30 and 70 (`NCDMORT3070`) for females in the United States over time. Make sure to add a title and informative axis labels to your plot.

```{r}
mortality_usa_df = ncd_data %>% 
  filter(sex == "Female",
         indicator_abbr == "NCDMORT3070",
         country == "United States of America")

ggplot(..., aes(...)) +
  ... 
```

# Part 3: Plotting Groups

## Question 6 (5 points)

Using a line plot, visualize the estimates and confidence intervals (using `geom_ribbon()`) of underweight (`NCD_BMI_18A`) and obesity prevalence (`NCD_BMI_30A`) for females in the United States over time. Make sure to add a title and informative axis labels to your plot.

Hint: set `alpha = 0.2` inside `geom_ribbon()` to make the fill of the intervals transparent

```{r}
bmi_usa_df = ncd_data %>% 
  filter(sex == "Female",
         indicator_abbr %in% c("NCD_BMI_18A", "NCD_BMI_30A"), 
         country == "United States of America")

ggplot(..., aes(...)) + 
  ... 
```

## Challenge Question 1 (5 points)

Plot the estimates and confidence intervals of the prevalence of elevated blood pressure (`BP_04`) by sex in France, Germany, Spain, and Portugal in 2015. Plot estimates on the x-axis and countries on the y-axis. Each country should have two corresponding points + error bars that are plotted next to each other, one for prevalence among females and one prevalence among males. Color plot elements for males vs females in different colors.

Hint: look into how `position = position_dodge(...)` can be used to avoid overlapping elements of your plot

```{r}
drinking_west_europe_df = ncd_data %>% 
  filter(indicator_abbr == "BP_04", 
         year == 2015,
         country %in% c("France", "Germany", "Spain", "Portugal"))

ggplot(..., aes(...)) + 
  ... 
```

## Challenge Question 2 (5 points)

Using a line plot, visualize the estimates and confidence intervals (using `geom_ribbon()`) of the prevalence of tobacco use (`M_Est_tob_curr`) by sex in Vietnam, Cambodia, Laos, and Thailand over time. Use `facet_wrap` to create a subplot for each country. Within each facet, create a line plot with year on the x-axis, tobacco use prevalence on the y-axis, and use different colored lines for each sex. Make sure to add a title and informative axis labels to your plot.

```{r}
se_asia_hypertension_df = ncd_data %>% 
  filter(indicator_abbr == "M_Est_tob_curr", 
         country %in% c("Viet Nam", "Lao People's Democratic Republic", 
                        "Cambodia", "Thailand"))

ggplot(...) +
  ... 
```

# Submission Instructions

Knit this R notebook as a PDF and submit it to the corresponding Gradescope assignment. Check the file to make sure that all your code and its output is displayed.

Once you successfully submit, you will be asked to match each question to its respective page on your submission. **Do not skip this step.**
