---
title: 'Functions and Packages'
output: html_document
---

In this chapter, you'll learn all about functions in R. This includes building your own functions, using R's built-in functions, and loading packages that contain helpful tools.

# Functions

## Function? I sure hope it does.

Functions allow us to take an input, run some code, and send back an output. In fact, if you've ever taken a math class before, you've already seen functions!

$$
f(x) = 2x
$$

Look familiar? This function is very simple. It takes an input $x$, multiplies it by two, and gives back the answer. You can run it over and over again for different values:

$$
f(1)=2*1=2\\
f(-10)=2*-10=-20\\
f(0)=2*0=0
$$

R lets us do the same exact thing. We can provide instructions inside a function that we use later for different inputs. R uses `function(){}` to define these inputs and instructions. The inputs go inside the `()` and the instructions go inside the `{}`. In this case, our input `x` goes inside the parentheses, and our instructions `2*x` goes inside the curly brackets. We give the function the name `times_two` so we can reference it later. So if we want to turn our math function into an R function called `times_two`, we can say:

```{r}
times_two = function(x){
  2*x
}
```

Run this code cell. R creates a function called `times_two` that you can use in later code cells in this notebook. We can **call** the function using its name followed by our input in parenthesis. For `times_two`, all we have to do is give `x` a value:

```{r}
times_two(x=1)
times_two(x=-10)
times_two(x=0)
```

Each time we run the function, it **returns** back the last thing it calculated. For this example, it only calculates `2*x`, so that is what it returns.

We can also use a `return()` statement to explicitly state what we want the function output to be. A `return()` statement tells R that we want to exit the function, and any code after it won't be run.

```{r}
multiply = function(x, y){
  return(x*y)
  x = x + 1 # This code will never be reached because return exits the function
  x
}

multiply(x=4, y=10)
```

### Practice Problem 1

Write your own function called `times_four_plus_one` that takes one input `x`, multiplies it by four, and adds one. Then, call your function three times with different inputs to make sure it works right.

```{r}
...
```

::: {.callout-tip collapse="true" title="Solution"}
```{r}
times_four_plus_one = function(x){
  4*x + 1
}
times_four_plus_one(1)
times_four_plus_one(-10)
times_four_plus_one(0)
```
:::

## Built-in Functions

R has a bunch of built-in functions that help us get our work done. In fact, you have been using functions this whole time! `data.frame()` is a useful function that we learned about in the Data Structures chapter. And in Data Types, we learned about a bunch of math functions like `exp()`, `log()`, and even `abs()` which gives us the absolute value!

An especially useful built-in function is `seq()`. This function allows you to generate a sequence of numbers. Now, we already learned that the colon `:` can be used to generate numbers in order. `seq()` expands on this by taking three inputs: `from=`, `to=`, and `by=`. `from` tells it our starting number, `to` tells it our maximum number, and `by` tells it the gaps between the numbers. If we want the numbers 1 through 10, we can just say `1:10` or `seq(from=1, to=10, by=1)`. However, if we want every odd number from 1 to 10, we cannot use the `:` and have to use `seq(from=1, to=10, by=2)`.

```{r}
# 1 through 10
1:10
seq(1, 10)
# odd numbers 1 through 10
seq(1, 10, 2)
```

### Practice Problem 2

Now you use `seq()` to generate all multiples of 0.2 from 0 to 5 (e.g. 0, 0.2, 0.4, ... , 4.8, 5).

```{r}
# Write your code here
```

::: {.callout-tip collapse="true" title="Solution"}
```{r}
seq(from=0, to=5, by=0.2)
```
:::

## Getting Help

R has a lot of functions built in, so it is almost impossible to remember all of them. Instead, we have the `?` operator. When you write this operator before a function name in R, it will pop up some useful information.

```{r}
?sample
```

When we run this line, a little window pops up on the right side describing what `sample()` does. We learn that it takes a sample of size `size` from the elements of `x`. So let's randomly pick 3 numbers between 1 and 10.

```{r}
sample(x=1:10, size=3)
```

There is also an optional input `replace` that is normally equal to `FALSE`. Further down under **Arguments**, we learn that `replace` tells it "should sampling be with replacement?". In other words, should the same element be able to be picked multiple times? If we want that to be yes, we can tell the function that. Try running this code multiple times until you get a repeated number.

```{r}
sample(x=1:10, size=3, replace=TRUE)
```

The help page on the right side also has examples down at the bottom to help us. You can also always use Google to help you figure out how to use these functions, but the `?` operator works even without Internet!

### Practice Problem 3

Use `?runif` (and the Internet if you need) to figure out how to get a random number between -1 and 2.

```{r}
# Write your code here
```

::: {.callout-tip collapse="true" title="Solution"}
```{r}
runif(1, min=-1, max=2)
```
:::

## Any Number of Arguments

Sometimes, functions will have an input that just says `…`. For example:

```{r}
?max()
```

The ellipsis means that `max()` can accept any number of arguments and give you the maximum of them.

```{r}
max(1, 5, 22)
```

### Practice Problem 4

Try it out! The `max()` and `min()` functions will give you the maximum and minimum, respectively, of numbers or letters. They will accept any number of inputs, including a pre-made vector.

Try finding different minimums and maximums. What happens when you include an `NA` in the group of numbers? What can you do about that (hint: check `?max`?

```{r}
min()
max()

some_numbers = c(...)
max(some_numbers)
```

::: {.callout-tip collapse="true" title="Solution"}
```{r}
min(3, 1, 9, 2)
max(3, 1, 9, 2)
some_numbers = c(3, 1, 9, 2, NA)
max(some_numbers, na.rm=TRUE)
```
:::

# Control Statements

Sometimes we write code that we only want to be run sometimes, or maybe we want to run it many times. For those situations, we use control statements. R gives us ways to say "only run this code if this is true" or "run this code over and over again". These code statements will look something like this:

```         
type(control_statement){
  code_to_be_run
}
```

The `type` tells R how we want the code to be run (sometimes, many times, etc). The `control_statement` is code we write to say when the code should be run or how many times it should be run. The `code_to_be_run` is inside curly brackets `{}` and can be any code, including multiple lines of code.

## If-Else Statements

If-else statements give us a chance to decide what code should be run. They will look like this:

```{r}
if(conditional_statement){
  # If conditional_statement is TRUE, this code will be run
} else {
  # If conditional_statement is FALSE, this code will be run
}
```

We can use this statement to have code that is run in different situations. For example, if we want to tell the user if they try to do some math that doesn't work, we could use an if-else statement.

```{r}
x <- -2

if(x >= 0) {
  sqrt(x)
} else {
  print("You cannot take the square root of a negative number.")
}

x <- 625

if(x >= 0) {
  sqrt(x)
} else {
  print("You cannot take the square root of a negative number.")
}
```

You can also use the `if` part on its own if there is not a second step you want to do. You can also put multiple if-else statements together if there are multiple decisions you need to make. Feel free to run this code multiple times with different `x` values.

```{r}
x <- 6

# Only make one decision
# If it's greater than 10, say that. Otherwise do nothing.
if(x > 5) {
  print("x is greater than 5")
}

# Categorize x into whether it's over 10, 9-5, 5-3, or less than or equal to 2.
if (x > 10) { 
  print("x is greater than 10")
} else if (x > 5) {
  print("x is 9 to 5")
} else if (x > 2) {
  print("x is 5 to 3")
} else {
  print("x is less than or equal to 2")
}
```

Note than when a condition in an if-else series ends up being `TRUE`, all of the other `else`s are ignored. In the example above, `x > 5` is `TRUE`, so `x is 9 to 5` is printed. The everything after the `else` gets ignored. Even though the next `if (x >2)` would have also been `TRUE`, we don't see it print that out because it is after the `else`.

### Practice Problem 5

Complete the code below to write a series of conditional statements that compares the values of `x` and `y`. Try changing the values of `x` and `y` to check that your code works as expected

```{r}
x = 24
y = 10

if (x > y) {
  print("x is ... than the value of y")
} else if (...) {
  print("x is less than the value of y")
} else {
  print("x is equal to y")
}
```

::: {.callout-tip collapse="true" title="Solution"}
```{r}
if (x > y) {
  print("x is greater than the value of y")
} else if (x < y) {
  print("x is less than the value of y")
} else {
  print("x is equal to y")
}
```
:::

## Loops

Many times in research, we will need to run the same code over and over again, like if we need to calculate a value for every participant in a study. R gives us a way to do this using *for loops*. A typical for loop might look like this:

```{r}
numbers <- 1:5

# for(element in collection_of_elements)
for (number in numbers) {
  # Code that gets repeated
  print(paste("Number is now", number))
  print(number*5)
}
```

In the for loop, we specify how many times the code needs to be run by giving it a collection of things, the vector `numbers` in this case. We give that element a name, `number`, and write code using that variable. In this way, the first time the loop runs, `number` will be set to 1 because that is the first element of `numbers`. The code inside the curly brackets runs. Then `number` is set to 2 because it is the second element of `numbers`. The loop continues all the way through 5 and then stops because 5 is the last element in the vector `numbers`. When writing your own for loop, you use the format `for (variable in values)`. `values` needs to be a collection of something that you have already created. `variable` is a placeholder name that you can use inside the curly brackets of the for loop to do what you want to do.

Something we normally do inside for loops is add the result to a list of results so we can keep track of all of them. Saving your results is a lot better than just printing them out, especially when you have a lot of observations. Let's say we have ten people in a study, and we want to categorize them by age. If we have their ages stored in `ages`, we can use a for loop to add their categorized age to `age_categories`. Notice that the part labeled "Step 2" is *inside* the for loop, so it will add `new_age_category` to the end of `age_categories` every time the loop is run.

```{r}
ages = c(25, 8, 35, 87, 34, 15, 73, 34, 53, 2)

age_categories = c()
  
# Categorize the first age
for (age in ages) {
  # Step 1: Match age to the correct category
  if (age < 18) {
    new_age_category = "<18 years"
  } else if (age < 64) {
    new_age_category = "18-64 years"
  } else {
    new_age_category = ">=65 years"
  }
  
  # Step 2: Add age category to the vector
  age_categories = c(age_categories, new_age_category)
}

# Display age and age category side-by-side
cbind(ages, age_categories)
```

### Practice Problem 6

Complete the code below to categorize the stage of chronic kidney disease (CKD) based on estimated glomerular filtration rate (eGFR) values in `egfr_values`

eGRF values less than 15 correspond to Stage 5 CKD, values between 15 and 29 correspond to Stage 4, values between 30-44 correspond to Stage 3B, and values above 45 correspond to Stages 1-3A.

```{r}
egfr_values = c(24, 61, 8, 36, 98, 78, 47, 42, 20)
ckd_statuses = c()

for (... in ...){
  # Convert one egfr value to CKD stage
  if (... < 15) {
    stage = "5"
  } else if (... < 30){
    stage = "4"
  } else if (... < 45){
    stage = "3b"
  } else {
    stage = "1 to 3a"
  }
  # Add CKD stage to vector
  ckd_statuses = c(..., ...)
}

ckd_statuses
```

::: {.callout-tip collapse="true" title="Solution"}
```{r}
for (egfr in egfr_values){
  if (egfr < 15) {
    stage = "5"
  } else if (egfr < 30){
    stage = "4"
  } else if (egfr < 45){
    stage = "3b"
  } else {
    stage = "1 to 3a"
  }
  ckd_statuses = c(ckd_statuses, stage)
}
```
:::

# Packages

Writing functions sure is fun, but I don't want to have to write a function for everything in R. Luckily, many amazing data scientists have come before us, written useful functions, and stored them in packages. Packages are collections of useful functions you can use in your own code.

One of the most useful packages is called **tidyverse**. It includes many tools for cleaning, summarizing, and visualizing data. On this website, WebR **pre-installs and loads** the tidyverse packages you need for each page, so you can use them in the interactive code cells without any setup.

If you were running the same code in **RStudio on your own computer**, you would need two steps first:

1.  Install the package once with `install.packages()`
2.  Load it in each R session with `library()`

```{r}
# These steps are only needed in RStudio (or other local R environments).
# On this website, tidyverse is already available and running this code results in an error
install.packages("tidyverse")
library(tidyverse)
```

Once tidyverse is available, you can use its functions right away. For example, `if_else()` from the `dplyr` package (part of tidyverse) chooses between two values based on a condition, similar to we used `if` and `else` earlier in this chapter.

The code below uses `if_else()` to label measurements as `"adult"` or `"child"` based on if their age value is over 18:

```{r}
ages = c(12, 45, 67, 8)
if_else(ages >= 18, "adult", "child")
```

# Summary

In this tutorial, you learned how to write your own functions, use built-in R functions, and get help with `?`. You also practiced control statements—`if`/`else` for conditional code and `for` loops for repeating actions—and were introduced to packages, including tidyverse.

Next, you'll learn how to import datasets and explore data frames with `readr` and `dplyr`.

# Extra Practice

## Question 1
Write a simple function that takes a character input and prints (use `print()`) that input.

```{r}
# Write your function here
printer

# Check your function
printer(input = "Hello world")
printer(input = "Go trees!")
```
::: {.callout-tip collapse="true" title="Solution"}
```{r}
printer = function(input) {
  print(input)
}

printer(input = "Hello world")
printer(input = "Go trees!")
```
:::

## Question 2 

Write a function that takes two inputs, `base` and `exponent`, that returns the base raised to the exponent.

```{r}
# Fill in the function with the inputs and code needed to calculate
exponentiate <- function(...) {

}

# Check your function
exponentiate(base=2, exponent=5) # Should be 32
exponentiate(base=5, exponent=-1) # Should be 0.2
```
::: {.callout-tip collapse="true" title="Solution"}
```{r}
exponentiate <- function(base, exponent) {
  base^exponent
}

exponentiate(base=2, exponent=5)
exponentiate(base=5, exponent=-1)
```
:::

## Question 3

When using the built-in `log` function, taking the log of 0 outputs `-Inf` (negative infinity), and taking the log of a negative number outputs `NaN` (not a number).

```{r}
log(0)
log(-2)
```

Write a function called `smart_log` that takes any number as the input. If that number is zero or negative, return `NA`. Otherwise, return the log of that number.

```{r}
smart_log <- function(...){
  # If statement to check if the number is zero or negative that returns NA
  
  # Otherwise, return the log of the number
  
}

# Check your function
smart_log(num=10)
smart_log(num=0)
smart_log(num=-2)
```
::: {.callout-tip collapse="true" title="Solution"}
```{r}
smart_log <- function(num){
  if (num <= 0) {
    return(NA)
  }
  log(num)
}

smart_log(num=10)
smart_log(num=0)
smart_log(num=-2)
```
:::

## Question 4

Now that we have our function, we can use a for-loop to find the log of a bunch of numbers.

Write a for-loop that calculates and and saves the log of each number in a vector.

```{r}
# Calculate the log of these numbers using the function you made above
# Save the results to logs
numbers <- c(1, -10, 20, 50, 0)
logs <- c()

# Write your for-loop that calculates and saves the log of each number using 
# the function you made above
for(...) {
  # Use the smart_log function to get the log and save it to a variable
  
  # Add that variable to the logs list
  
}

# Check your results
logs
```
::: {.callout-tip collapse="true" title="Solution"}
```{r}
numbers <- c(1, -10, 20, 50, 0)
logs <- c()

for (num in numbers) {
  result = smart_log(num)
  logs = c(logs, result)
}

logs
```
:::

## Question 5

Remember this formula?

$$
x=\frac{-b\pm\sqrt{b^2-4ac}}{2a}
$$

It's the quadratic formula from algebra, which tells us where a parabola intercepts the $x$ axis. Write a function that takes three values, `a`, `b`, and `c`, and returns a vector with two values - the two places where the parabola crosses the $x$ axis. But, if `a` is zero, return a vector with two `NA` values because that is not a parabola.

Hint: There is no $\pm$ in R. You'll have to do those separately.

```{r}
# Write your function
get_parab_x_intercepts = ...

# Testing
get_parab_x_intercepts(a=2, b=4, c=-3) # Should be 0.5811 and -2.5811
get_parab_x_intercepts(a=-5, b=1, c=6) # Should be -1 and 1.2
get_parab_x_intercepts(a=0, b=10, c=2) # Should be NA and NA
```
::: {.callout-tip collapse="true" title="Solution"}
```{r}
get_parab_x_intercepts = function(a, b, c) {
  if (a == 0) {
    return(c(NA, NA))
  }
  x1 = (-b + sqrt(b^2 - 4*a*c)) / (2*a)
  x2 = (-b - sqrt(b^2 - 4*a*c)) / (2*a)
  c(x1, x2)
}

get_parab_x_intercepts(a=2, b=4, c=-3)
get_parab_x_intercepts(a=-5, b=1, c=6)
get_parab_x_intercepts(a=0, b=10, c=2)
```
:::

## Question 6

Use the `?` operator to figure out how to use the `seq()` function get every fifth number from 0 to 90 (i.e. 0, 5, 10, ..., 90). Use the same function to get every 0.2 from 2 to 5 (i.e. 2.0, 2.2, 2.4, ..., 5).

```{r}

```
::: {.callout-tip collapse="true" title="Solution"}
```{r}
seq(from=0, to=90, by=5)
seq(from=2, to=5, by=0.2)
```
:::
