---
title: "Final Project"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

Similar to Homework 4, your final will consist of a start-to-finish
analysis of pre-existing data. We will give you a structure and a
research question to examine. It is your responsibility to organize your
code, give appropriate variable names, and explain the output of your
code so that it could be understood by someone who does not know R.
Please make sure to make good use of comments, check your indenting, and
ensure no code runs off the page to get full style points!

We recommend that you complete this assignment in the **Visual** view of
the notebook.

# Background

Your final project uses data from [this study published in *Nature
Microbiology*](https://www.nature.com/articles/s41564-023-01543-3 "Household immunity and individual risk of infection with dengue virus in a prospective, longitudinal cohort study").

The study examines individual- and household-level risk factors of
dengue in Kamphaeng Phet province, Thailand. Dengue is a mosquito-borne
viral infection that is found in tropical climates. Symptoms of dengue
include fever, headache, rashes, and muscle and join pain; in extreme
cases, dengue can also cause bleeding, low levels of blood platelets,
and low blood pressure. However, most dengue infections are asymptomatic
or only result in mild symptoms.

The data for your final are based off of what the authors have
published, but we have done some pre-processing to simplify the
analysis. **You will not be able directly replicate the paper's
results**, so do not worry if any of your data summaries or inferences
are different from what's reported in the paper.

You will work with two complementary datasets in this project. These two
datasets are linked by a household ID.

-   `dengue_individual_data.csv` contains individual-level
    characteristics and infection status

-   `dengue_household_data.csv` contains household characteristics

# Loading packages and data

Run the cell below to load the required packages

```{r}
library(tidyverse)
library(here)
```

## Individual data

```{r}
individual_data = read_csv(here("data", "dengue_individual_data.csv"))
glimpse(individual_data)
```

**Data dictionary:**

| Variable Name        | Description                                                                              |
|--------------------------|----------------------------------------------|
| `ID`                 | Individual identifier                                                                    |
| `hh_ID`              | Household identifier                                                                     |
| `sex`                | Biological sex                                                                           |
| `n_symp_infections`  | Count of times this individual has had a symptomatic dengue infection during the study   |
| `n_asymp_infections` | Count of times this individual has had an asymptomatic dengue infection during the study |

## Household data

```{r}
household_data = read_csv(here("data", "dengue_household_data.csv"))
glimpse(household_data)
```

**Data dictionary:**

| Variable Name          | Description                                                                                                                                                                                                                                                                                                       |
|-------------------------|-----------------------------------------------|
| `hh_ID`                | Household identifier                                                                                                                                                                                                                                                                                              |
| `door_screens`         | 1, if the household has a screen door. 0, otherwise                                                                                                                                                                                                                                                               |
| `house_type`           | One of three household types. "Poles" indicates the house is built on stilts, as homes on floodplains typically are. "Single" indicates the house is a single-family home separate from other homes. "Townhouse" indicates the house if part of a multifamily unit where multiple homes are physically connected. |
| `concrete_house`       | 1, if the house was built from concrete. 0, otherwise                                                                                                                                                                                                                                                             |
| `water_supply_by_pipe` | 1, if the house is supplied water by pipes. 0, if the house receives water some other way                                                                                                                                                                                                                         |
| `garbage_management`   | One of three methods for getting rid of household waste. "Burn" if it is burned by the residents. "Car" if it is disposed of by a vehicle that collects waste. "Dump" if it is brought to a waste disposal facility by the residents.                                                                             |
| `n_toilets_outside`    | Count of the outdoor toilets available to the household residents                                                                                                                                                                                                                                                 |

# Research Question

For your final, you will examine this research question "Are household
factors associated with dengue infection?" Follow the steps below to
complete the analysis. Make sure to document your code using plain
language so that a researcher who does not know R could understand.

# Data Exploration

## Question 1 (2 points)

Explore the distribution of at least one continuous variable and one
categorical/binary variable in the datasets using both summary
statistics and a exploratory plots.

```{r}

```

## Question 2 (2 points)

These is one column in `household_data` with missing/unknown values. How
are the missing values coded? Identify the column and find the
percentage of measurements that have a missing value for that variable.

```{r}

```

# Data Cleaning

## Question 3 (1 point)

Create a copy of `individual_data` called `individual_data_cleaned` and
a copy of `household_data` called `household_data_cleaned`. You will
save the outputs of all your data cleaning and processing to these
"cleaned" datasets to avoid overwriting the original datasets.

```{r}
individual_data_cleaned = ...
household_data_cleaned = ...
```

## Question 4 (2 points)

For the column in `household_data_cleaned` you identified in Question 2,
recode missing or unknown values as `NA` and convert the remaining
values to numerics.

```{r}

```

## Question 5 (4 points)

Add the following columns to `individual_data_cleaned`:

-   `age_cat`, that is "adult" for people 18 or older and "child" for
    people under 18.

-   `n_any_infections`, the sum of symptomatic and asymptomatic
    infections for each person.

-   `any_infections`, that is `TRUE` if the individual was ever infected
    with symptomatic or asymptomatic dengue.

```{r}

```

## Challenge Question 1 (3 points)

Add a `household_size` variable to `household_data_cleaned` that counts
the number of individuals in each household.

*Hint:* Use the individual dataset to calculate the number of people in
each household, then merge the data into `household_data_cleaned`

```{r}

```

## Question 6 (3 points)

Join the cleaned individual and household level data to create a dataset
called `merged_dengue_data`. The resulting data frame should contain 1
row per individual, with columns for both individual- and
household-level variables for that person.

```{r}
merged_dengue_data = ...
```

# Descriptive Analysis

## Question 7 (3 points)

Fill-in the values for Table 1. The "Total" column should describe
characteristics for the full sample, while "No Infection" and "Any
Infection" describe characteristics for subgroups by infection status.

The cells with "---" in them do not need to be filled in. The percents
should be column percents (ex. % no infection males + % no infection
females = 100%)

You should find all quantities indicated in the left column, including
sample sizes, percentages, means, and SDs, using R code (no hand
calculations). Clearly label and document your code so it's clear how
you calculated each value.

```{r}

```

| Table 1                  | Total | No Infection | Any Infection |
|--------------------------|:-----:|:------------:|:-------------:|
| Number of individuals, N |       |              |               |
| Age, mean (SD)           |       |              |               |
| Sex, N (%)               |  ---  |     ---      |      ---      |
| \- Female                |       |              |               |
| \- Male                  |       |              |               |
| Concrete House, N (%)    |  ---  |     ---      |      ---      |
| \- Yes                   |       |              |               |
| \- No                    |       |              |               |
| House Type, N (%)        |  ---  |     ---      |      ---      |
| \- Poles                 |       |              |               |
| \- Single                |       |              |               |
| \- Townhouse             |       |              |               |
| Water Supply, N (%)      |  ---  |     ---      |      ---      |
| \- Pipes                 |       |              |               |
| \- Other                 |       |              |               |
| \- Missing               |       |              |               |

## Challenge Question 2 (3 points)

Complete the rest of Table 1.

You should find all quantities, including sample sizes, percentages,
means, and SDs, using R code (no hand calculations). Clearly label and
document your code so it's clear how you calculated each value.

```{r}

```

| Table 1                      | Total | No Infection | Any Infection |
|------------------------------|:-----:|:------------:|:-------------:|
| Screen Door, N (%)           |  ---  |     ---      |      ---      |
| \- Yes                       |       |              |               |
| \- No                        |       |              |               |
| Garbage Management, N (%)    |  ---  |     ---      |      ---      |
| \- Burn                      |       |              |               |
| \- Car                       |       |              |               |
| \- Dump                      |       |              |               |
| Household Size, Median (IQR) |       |              |               |

## Question 8 (3 points)

Calculate the odds ratio for being infected with dengue
(`any_infections`) by whether the house is made of concrete.

Do not "hard code" this calculation by copying exact numbers - use R
code to first find the number of infections in concrete vs non-concrete
homes. Your code for the odds ratio calculation should still work even
if the underlying data changes.

```{r}
n_infections_concrete = ...
n_no_infections_concrete = ...
odds_concrete = ...

n_infections_nonconcrete = ...
n_no_infections_nonconcrete = ...
odds_nonconcrete = ...

concrete_OR = ... 
concrete_OR
```

## Challenge Question 3 (3 points)

Calculate the odds ratio for being infected with dengue by adults vs
children.

Do not "hard code" this calculation by copying exact numbers - use R
code to first find the number of infections in adults vs children. Your
code for the odds ratio calculation should still be valid even if the
underlying data changes

```{r}

```

# Statistical Inference

## Question 9 (2 points)

Run a chi-square test to assess if there is a difference in any dengue
infection between males and females.

Interpret your model output below your code. (In what group were
infections higher? Was the difference statistically significant?)

```{r}

```

*Interpretation:* ...

## Question 10 (3 points)

Use a linear model to assess how the number of outdoor toilets affects
the number of dengue infections. Get a 95% confidence interval for your
estimate.

Interpret your model output below your code. (For each addition outdoor
toilet, what is the expected change to the number of dengue infections?
Is the effect statistically significant?)

```{r}

```

*Interpretation:* ...

## Question 11 (3 points)

Use a logistic model to estimate the odds ratio and 95% confidence
interval of any dengue infection associated with screen doors.

Interpret your model output below your code. (How do the odds of
infection change for homes with vs without screen doors? Is this a
statistically significant effect?)

```{r}

```

*Interpretation:* ...

## Question 12 (3 points)

Extend the model you created in Question 5 to estimate the odds ratio
and 95% confidence interval of any dengue infection associated with
screen doors, controlling for house type and garbage management.

Interpret your model output below your code. (How do the odds of
infection change for homes with vs without screen doors, when
controlling for other factors? Is this a statistically significant
effect?)

```{r}

```

*Interpretation:* ...

## Challenge Question 4 (5 points)

Fill in the skeleton code to produce a bootstrapped estimate of the risk
ratio of any dengue infection for children vs adults using 100
bootstrapped samples.

*Hint:* Reference the "Optional Challenge: Bootstrapping" section of the
Week 7 tutorial for help!

```{r}
# Create a vector to store the risk ratios
...

# Number of bootstrap samples to take
bootstrap_n = ...

# Create a for loop to run through the number of times to bootstrap calculating
#   a the risk ratio of any infection for children versus adults from a random 
#   sample of the data and saving it to your risk ratio vector
for (...) {
  # Draw bootstrap sample WITH replacement (use slice_sample())
  bootstrap_sample = ...
  
  # Calculate the risk of any infection in children
  risk_children = ...
  
  # Calculate the risk of any infection in adults
  risk_adult = ...
  
  # Add risk ratio to your vector of risk ratios
  risk_ratio = risk_children/risk_adult
  ... = c(...)
}

# Take the mean of your risk ratios to get your estimate
...

# Get the 95% confidence interval from your risk ratio vector
quantile(..., probs=c(0.025,0.975))
```

# Visualizations

***You must use `ggplot` for this section.*** **Make sure each plot has
an informative title and axes names.**

## Question 13 (3 points)

Make a scatterplot of the number of any infections by age. Place age on
the x-axis and the number of any infections on the y-axis.

```{r}

```

## Question 14 (4 points)

Make a boxplot of the number of any dengue infections by house type.
Place house type of the x-axis and infections on the y-axis. Order the
house type from left to right as "Single", "Townhouse", and "Poles".

```{r}

```

# Submission Instructions

Knit the completed document as a PDF and submit it on Gradescope by the
deadline. For students taking the course for 1 unit, you may need to
delete the skeleton code under the challenge questions for your code to
knit properly. Please double check the knitted PDF to make sure all of
your code and comments are fully visible and do not run off the screen.
